The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-01-09 Paper.Rmd
further edits to Q1 materials

Very generally, all branches of ecology and related areas study relationships among environmental and biological 
variables. However, commonly used correlation- and regression-based 
approaches to studying 
such relationships are limited - it is well understood that common statistical
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value;  
Fig. \ref{fig_cop_pedag1}A-B shows two such datasets. The variables of panel A 
(respectively, B) are principally 
related in the left (respectively, right) portions of their distributions, an asymmetric pattern of dependence in the distribution tails that may
have ecological significance, as discussed below, but is not captured by 
correlation. Panels C and D 
were obtained from A and B through monotonic transformation 
(Appendix \ref{SM-pedagog1_details} for details),
<!--***DAN: Dan to go back and write the appendix-->
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of the numerous metrics for 
studying relationships between variables, other commonly used metrics also 
provide partial information only.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-F, I, J) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. Panels C and E (respectively, D and F) were 
obtained from A (respectively, B) via monotonic functions that transformed the standard-
normal marginal distributions of A and B to gamma (C, D) or 
uniform (E, F) marginals (details in Appendix \ref{SM-pedagog1_details}). (G, H) Contours 
of log-scale density functions of Clayton (G) and survival-Clayton 
(H) copulas with parameter 5 (Appendix \ref{SM-pedagog1_details}). (I,J) Reflections
of E, F, respectively, about a vertical axis. Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}
<!--***DAN: Dan to go back and add the appendices with details as it says. Make sure
to mention the shape and scale params of the gamma marginals in C, D were 2-->

Well-developed statistical tools do exist, and are 
applied widely in other fields, that provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence between the variables). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
or copula structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking serves to uniformize the 
marginal distributions of the underlying random variables, isolating the 
depedence structure only.
Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cummulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence or copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. Strictly speaking, 
the copula is actually defined as the bivariate 
cdf of $(F_X(X),F_Y(Y))$ rather than as this random variable itself, 
but this makes little conceptual difference. See @nelsen2006_copula and @joe2014_dependence
for mathematical details. An important theorem of A. Sklar [@sklar1959_theorem] 
states that copulas give a unique and complete 
representation of the dependence
between variables, in the sense that the joint distribution of $(X,Y)$ 
can be reconstructed completely from the 
copula and the marginal distributions. 
Copula theory is also well developed for
multivariate distributions.

A principal benefit of a copula approach to ecological data is 
that such an approach can sensitively model and detect
tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of two variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, E);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, F). Datasets of the same 
corelation can have a range of tail dependence 
characteristics (Fig. \ref{fig_cop_pedag}), which can be quantified (Methods).

\begin{figure}
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{(A-C) Data generated using the BB1 copula (Methods) with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths (Methods), but the same 
Kendall correlation (KC) (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column of panels.\label{fig_cop_pedag}}
\end{center}
\end{figure}

One way we may expect a copula approach 
to better reflect ecological relationships has to do with Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, as well
as on other factors, data on soil nitrogen and resulting growth rates may be 
expected to look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a strong
relationship, only when it is limiting, in the left portion of the N and 
plant-growth-rate distributions. 
Nitrogen will be less related or unrelated to growth rates in the right tails of 
the distributions, as pictured in Fig. \ref{fig_cop_pedag1}A and E. The portion
of the distributions that dominate the overall association is visible in the copula
approach, but may not be reflected by correlation and other standard techniques.

A second reason we may expect a copula-based approach to benefit 
ecological research is that such an approach may help illuminate 
causal mechanisms by which variables are related. For instance, if two species
X and Y are strong competitors, we may expect abundances of the two species
across quadrats or through time within a single quadrat to be negatively related, as in 
Fig. \ref{fig_cop_pedag1}I-J. If X is plotted on the horizontal axis
and Y is plotted on the vertical, then the plot in Fig. \ref{fig_cop_pedag1}I 
suggests X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed 
by X; whereas when X is rare, Y
may be abundant or may also be relatively rare due to 
limiting factors other than X. 
Alternatively, Fig. \ref{fig_cop_pedag1}J suggests that Y
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
detail may help support or refute alternative mechanistic and causal hypotheses about 
the relationship between the variables.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of catches from 
local areas. We will explore how details, captured by copula methods, of the 
nature of the dependence between local quantities can influence the spatial 
mean or sum, and how this may influence human concerns and 
higher organizational levels in 
ecology. Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks tend 
to occur together, creating regional epidemics. If populations
of a crop or other resource species are strongly 
related in their left tails, then 
episodes of local scarcity occur together, creating regional 
shortages. On the other hand left-tail dependence in the pest species 
and right-tail dependence in the crop, even if overall correlation were 
unchanged, would have much more benign effects.

We approached our overall goal of illustrating the potential of 
copula approaches for understanding dependences in ecology by addressing the 
folling specific questions. (Q1) Do datasets in ecology and related fields 
have copula structure distinct from a Gaussian/normal distribution (here called 
non-normal copula structure)? Do they show tail dependence distinct from that 
of a normal distribution, and in particular do they show asymmetric tail 
dependence? (Q2) If so, what are some possible causes/mechanisms of this 
non-normal copula structure? And (Q3) what are the consequences of non-normal 
copula structure and tail dependence for our understanding of biology and for 
applications? To address Q1, we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data. We selected these datasets to span multiple 
sub-fields and organizational levels of ecology, in an attempt to be 
reasonably representative of the field as a whole. The second section of the 
paper introduces the data, and the third and fourth sections use the data to 
address Q1. The fifth and sixth sections, which address Q2, use simple models to 
explore mechanisms. The seventh and eigth sections, which address Q3, use both
the data section two as well as models to explore a variety of consequences
of copula structure and tail dependence for applications and for our understanding
of ecology. The ninth section is the Discussion. 
Copula approaches have been used to great benefit in neuroscience [@onken2009], 
bioinformatics 
[@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], finance, and other fields. 
To our knowledge, these approaches have been used seldom or never for ecological 
research. We argue that the benefits to ecological research of a complete statistical 
desription of dependence will be substantial, and we hope this paper stimulates the
use of copula approaches across the discipline.

@@

We addressed Q1 partly via a model selection procedure in which the normal copula
and a variety of alternative copulas were fit to 
our ecological datasets, and fits of the various copula models 
were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). To make the results interpretable
to readers unfamiliar with copulas, we first introduce the normal copula and 
some other copulas we used; and we operationally introduce measures of tail dependence
and asymmetry of tail dependence. 
Conceptually, a bivariate copula is the same as a probability density function 
on the unit square with uniform maginals. The normal copula family is a one-parameter
family of copulas, with parameter corresponding to the covariance of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag3}A-B for an example).
Left- and right-tail dependence are $0$ for the normal copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measure correspond to dependence stronger than that of a normal
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}G-H) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}G) and $0$ upper-tail dependence; the 
survival-Clayton, which is a rotation of the Clayton by $180^\circ$, has 
positive upper-tail dependence ($0.87$ in Fig. \ref{fig_cop_pedag1}H)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several one- and two-parameter families of copulas exhibiting a wide variety of 
positive lower-tail dependence values 
(Clayton, survival-Gumbel, survival-Joe, survival-BB6, henceforth
abbreviated, respectively, C, SG, SJ, SBB6) and upper-tail dependence values
(survival-Clayton, Gumbel, Joe, BB6, henceforth abbreviated, respectively, 
SC, G, J, BB6). We also considered families that could exhibit positive upper- or lower-tail
dependence, or both, depending on parameters (BB1, survival-BB1, BB7, survival-BB7, henceforth 
abbreviated, respectively, BB1, SBB1, BB7, SBB7). 
If LT and UT are used to denote 
our two measures of tail dependence, their difference is a measure of asymmetry of 
tail dependence. We also used two families that were asymmetric in their non-tail parts, 
but that had zero tail dependence (BB8, survival-BB8, abbreviated BB8 and SBB8).
In addition to the normal copula, we used another symmetric copula,
the Frank copula (abbreviated F; Fig. \ref{fig_cop_pedag3}C-D). 
Copulas are implemented in the `copula` and `VineCopula` packages for the 
R statistical computing language.
Additional details are in Appendix \ref{SM-copula_details} and in introductory texts 
on copulas [@nelsen2006_copula ; @joe2014_dependence] and in `VineCopula` documentation.
Visualizations of all the copulas we used are in Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig3.pdf}
\caption{Log-transformed pdfs (A, C) and samples from (B, D) example copulas.
Shown are the normal copula with parameter $0.8$ and the Frank copula with 
parameter $10$. See also Fig. \ref{fig_cop_pedag1}E-H for the Clayton and  
survival-Clayton copulas with parameter 5, and Appendix \ref{SM-copula_details}
for additional details. \label{fig_cop_pedag3}}
\end{center}
\end{figure}

Question Q1 was answered principally via analysis of the
data using a model selection approach. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. Ranks were $1$ for the smallest element in a set. 
Second, we did a test of the independence of the $u_i$ and $v_i$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. For keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

To test for asymmetry of dependence in upper and lower portions of distributions,
we used differences of upper and lower statistics.
The difference $\cor_l-\cor_u$ is expected to be positive if 
dependence in the left halves of the distributions is stronger 
than dependence in their right halves; likewise for $\Ps_l-\Ps_u$.
The difference $\Dsq_u-\Dsq_l$ (note the opposite order), is also expected to be 
positive under the same conditions. These statistics were used for smaller datasets.
For large datasets, we used $\cor_{l_b,u_b}-\cor_{1-u_b,1-l_b}$, 
$\Ps_{l_b,u_b}-\Ps_{1-u_b,1-l_b}$ and
$\Dsq_{1-u_b,1-l_b}-\Dsq_{l_b,u_b}$ with $l_b=0$ and $u_b$ close to $0$. 
How close depended on $n$, and was chosen to ensure that sample variation 
was not unduly large. We tested our statistics by testing whether 
$\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed the asymmetry of 
tail dependence known to be present in data simulated from various copulas 
(Appendix \ref{SM-Test_npa_stat}).

>>>>

Very generally, all branches of ecology and related areas study relationships among environmental and biological 
variables. However, commonly used correlation- and regression-based 
approaches to studying 
such relationships are limited - it is well understood that common statistical
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value 
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, an asymmetric 
pattern of dependence in the distribution tails that may
have ecological significance, as discussed below, but is not captured by 
correlation. Panels C and D 
were obtained from A and B through monotonic transformation 
(Appendix \ref{SM-pedagog1_details} for details),
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of the numerous metrics for 
studying relationships between variables, other commonly used metrics also 
provide partial information only.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-F, I, J) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. Panels C and E (respectively, D and F) were 
obtained from A (respectively, B) via monotonic functions that transformed the 
standard-normal marginal distributions of A and B to gamma (C, D) or 
uniform (E, F) marginals (details in Appendix \ref{SM-pedagog1_details}). (G, H) Contours 
of log-scale density functions of Clayton (G) and survival-Clayton 
(H) copulas with parameter 5 (see section on methods for Q1). (I,J) Reflections
of E, F, respectively, about a vertical axis. Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed statistical tools do exist, and are 
applied widely in other fields, that provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence between the variables). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
or copula structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking serves to uniformize the 
marginal distributions of the underlying random variables, isolating the 
depedence structure only.
Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cummulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence or copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. Strictly speaking, 
the copula is actually defined as the bivariate 
cdf of $(F_X(X),F_Y(Y))$ rather than as this random variable itself, 
but this makes little conceptual difference. See @nelsen2006_copula and @joe2014_dependence
for mathematical details. An important theorem of A. Sklar [@sklar1959_theorem] 
states that copulas give a unique and complete 
representation of the dependence
between variables, in the sense that the joint distribution of $(X,Y)$ 
can be reconstructed completely from the 
copula and the marginal distributions. 
Copula theory is also well developed for
multivariate distributions.

A principal benefit of a copula approach to ecological data is 
that such an approach can sensitively model and detect
tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of two variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, E);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, F). Datasets of the same 
corelation can have a range of tail dependence 
characteristics (Fig. \ref{fig_cop_pedag}), which can be quantified (see section on
methods for Q1).

\begin{figure}
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption{(A-C) Data generated using the BB1 copula (see section on methods for Q1) 
with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths (methods for Q1), but the same 
Kendall correlation (K) (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column of panels.\label{fig_cop_pedag}}
\end{center}
\end{figure}
<!--***DAN: Shya, I changed the abbreviation for Kendall Correlation from KC to K in the 
caption, please change in the code that generates the figures, as well-->

One way we may expect a copula approach 
to better reflect ecological relationships has to do with Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, as well
as on other factors, data on soil nitrogen and resulting growth rates may be 
expected to look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a strong
relationship, only when it is limiting, in the left portion of the N and 
plant-growth-rate distributions. 
Nitrogen will be less related or unrelated to growth rates in the right tails of 
the distributions, as pictured in Fig. \ref{fig_cop_pedag1}A and E. The portion
of the distributions that dominate the overall association is visible in the copula
approach, but may not be reflected by correlation coefficients and other standard techniques.

A second reason we may expect a copula-based approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms through which variables are related. For instance, if two species
X and Y are strong competitors, we may expect abundances of the two species
across quadrats or through time within a single quadrat to be negatively related, as in 
Fig. \ref{fig_cop_pedag1}I-J. If X is plotted on the horizontal axis
and Y is plotted on the vertical, then the plot in Fig. \ref{fig_cop_pedag1}I 
suggests X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed 
by X; whereas when X is rare, Y
may be abundant or may also be relatively rare due to 
limiting factors other than X. 
Alternatively, Fig. \ref{fig_cop_pedag1}J suggests that Y
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
detail may help support or refute alternative mechanistic and causal hypotheses about 
the relationship between the variables.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This process is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of catches from 
local areas. We will explore how details, captured by copula methods, of the 
nature of the dependence between local quantities can influence the spatial 
mean or sum, and how this may influence human concerns and 
higher organizational levels in 
ecology. Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks tend 
to occur together, creating regional epidemics. If populations
of a crop or other resource species are strongly 
related in their left tails, then 
episodes of local scarcity occur together, creating regional 
shortages. On the other hand left-tail dependence in the pest species 
or right-tail dependence in the crop, even if overall correlation were 
unchanged, would have much more benign effects.

We approached our overall goal of exploring the potential of 
copula approaches for understanding dependences in ecology by addressing the 
folling specific questions. (Q1) Do datasets in ecology and related fields 
have copula structure distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure)? Do they show tail dependence distinct from that 
of a normal distribution, and in particular do they show asymmetric tail 
dependence? (Q2) If so, what are some possible causes/mechanisms of this 
non-normal copula structure? And (Q3) what are the consequences of non-normal 
copula structure and tail dependence for our understanding of biology and for 
applications? To address Q1, we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data. We selected these datasets to span multiple 
sub-fields and organizational levels of ecology, in an attempt to be 
reasonably representative of the field as a whole. The second section of the 
paper introduces the data, and the third and fourth sections use the data to 
address Q1. The fifth and sixth sections, which address Q2, use simple models to 
explore mechanisms. The seventh and eigth sections, which address Q3, use both
the data of section two as well as models to explore a variety of consequences
of copula structure and tail dependence for applications and for our understanding
of ecology. The ninth section is the Discussion. 
Copula approaches have been used to great benefit in neuroscience [@onken2009], 
bioinformatics 
[@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], finance, and other fields. 
To our knowledge, these approaches have been used seldom or never for ecological 
research. We argue that the benefits to ecological research of a complete statistical 
desription of dependence will be substantial, and we hope this paper helps stimulate the
fruitful use of copula approaches across the discipline.

@@

We addressed Q1 partly via a model selection procedure in which the normal copula
and a variety of alternative copulas were fit to 
our ecological datasets, and fits of the various copula models 
were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). To make the results interpretable
to readers unfamiliar with copulas, we first introduce the normal copula and 
some other copulas we used; and we operationally introduce measures of tail dependence
and asymmetry of tail dependence. 
Conceptually, a bivariate copula is essentially a probability density function 
on the unit square with uniform maginals. The normal copula family is a one-parameter
family of copulas, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag3}A-B for an example).
Left- and right-tail dependence are $0$ for the normal copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measure correspond to dependence stronger than that of a normal
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}G-H) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}G) and $0$ upper-tail dependence; the 
survival-Clayton, which is a rotation of the Clayton by $180^\circ$, has 
positive upper-tail dependence ($0.87$ in Fig. \ref{fig_cop_pedag1}H)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several one- and two-parameter families of copulas exhibiting a wide variety of 
positive lower-tail dependence values 
(Clayton, survival-Gumbel, survival-Joe, survival-BB6, henceforth
abbreviated, respectively, C, SG, SJ, SBB6) and upper-tail dependence values
(survival-Clayton, Gumbel, Joe, BB6, henceforth abbreviated, respectively, 
SC, G, J, BB6). We also considered families that could exhibit positive upper- or lower-tail
dependence, or both, depending on parameters (BB1, survival-BB1, BB7, survival-BB7, henceforth 
abbreviated, respectively, BB1, SBB1, BB7, SBB7). 
If LT and UT are used to denote 
our two measures of tail dependence, their difference is a measure of asymmetry of 
tail dependence. We also used two families that were asymmetric in their non-tail parts, 
but that had zero tail dependence (BB8, survival-BB8, abbreviated BB8 and SBB8).
In addition to the normal copula, we used another symmetric copula,
the Frank copula (abbreviated F; Fig. \ref{fig_cop_pedag3}C-D). 
Copulas are implemented in the `copula` and `VineCopula` packages for the 
R statistical computing language.
Additional details are in introductory texts 
on copulas [@nelsen2006_copula ; @joe2014_dependence] and in `VineCopula` documentation.
Visualizations of all the copulas we used are in Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig3.pdf}
\caption{Log-transformed pdfs (A, C) and samples from (B, D) example copulas.
Shown are the normal copula with parameter $0.8$ and the Frank copula with 
parameter $10$. See also Fig. \ref{fig_cop_pedag1}E-H for the Clayton and  
survival-Clayton copulas with parameter 5. Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}
have visualizations of all the copulas we used, including more visualizations
of the Clayton and survival-Clayton copulas.\label{fig_cop_pedag3}}
\end{center}
\end{figure}

Question Q1 was answered principally via analysis of the
data using a model selection approach. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. Ranks were $1$ for the smallest element in a set. 
Second, we did a test of the independence of the $u_i$ and $v_i$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

To test for asymmetry of dependence in upper and lower portions of distributions,
we used differences of upper and lower statistics.
The difference $\cor_l-\cor_u$ is expected to be positive if 
dependence in the left halves of the distributions is stronger 
than dependence in their right halves; likewise for $\Ps_l-\Ps_u$.
The difference $\Dsq_u-\Dsq_l$ (note the opposite order), is also expected to be 
positive under the same conditions. These statistics were used for smaller datasets.
For large datasets, we used $\cor_{l_b,u_b}-\cor_{1-u_b,1-l_b}$, 
$\Ps_{l_b,u_b}-\Ps_{1-u_b,1-l_b}$ and
$\Dsq_{1-u_b,1-l_b}-\Dsq_{l_b,u_b}$ with $l_b=0$ and $u_b$ close to $0$. 
How close depended on $n$, and was chosen to ensure that sample variation 
was not unduly large. We tested our statistics by testing whether 
$\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed the asymmetry of 
tail dependence known to be present in data simulated from various copulas 
(Appendix \ref{SM-Test_npa_stat}, Figs \ref{SM-fig_stat_testing35}-\ref{SM-fig_stat_testing1000}).
